Strong lensing selection effects

### 2022-12-28 paper/seleff.tex
Ned and Anowar's comments

Throughout our work we fix the S\'{e}rsic index of the lenses to $n=4$, also known as a de Vaucouleurs model.
Although early-type galaxies are often described with a free S\'{e}rsic index, a de Vaucouleurs profile is able to reproduce their surface brightness distribution to a few percent in the radial range $1\rm{kpc} < R < 30\rm{kpc}$ \citep[see e.g.][]{Son20}, which is the region that is relevant for strong lensing.
Finally, we assume that the light distribution of the stellar component follows its mass distribution exactly. That is, we do not allow for the presence of gradients in the stellar mass-to-light ratio. This assumption is probably broken in real galaxies but, as we discuss in \Sref{sect:discuss}, does not impact our investigation.

@@

This result suggests that, to first approximation, lensing does not introduce a strong selection in source size.
This is expected, since gravitational lensing preserves surface brightness: a source that can be detected in the absence of lensing will produce images with the same surface brightness when lensed, which can be detected as well.
%When the source is extended, however, 
In order to classify a source as strongly lensed, however, we require that multiple images are observed. 
Only sources that lie within a well-defined region give rise to a strong lensing configuration.
If part of the source extends outside of this region, then only a fraction of its flux contributes to creating a set of strongly lensed images.
This lowers the signal-to-noise ratio of the multiple images compared to the point source case.
%The result is the decrease in $\crosssect$ at values of the source size smaller than the 
The result is that $\crosssect$ starts to decrease with increasing source size at values of $\theta_{\mathrm{e,s}}$ that are smaller than the no-lensing detection limit, as observed in \Fref{fig:extcs}.

@@

To determine the density profile of the halo we relied on a theoretically motivated model that takes into account the effect of baryons on the dark matter.
We assumed the dark matter distribution to be initially described by an NFW profile with a concentration\footnote{The concentration is the ratio between the halo virial radius and the scale radius $r_s$.} of five, then used the prescription of \citet{Cau++20} to model the response of the halo to the infall of baryons. % on the basis of the present stellar mass distribution. 
In the \citet{Cau++20} model the halo response is approximated with an analytical function that depends on the present stellar mass distribution, and typically results in a more concentrated and steeper density profile compared to the original NFW model. 
Finally, we fitted a gNFW profile to the surface mass density of the contracted halo. By doing this, we were able to fully describe the dark matter density profile with three parameters: $\mhalo$, $\gammadm$ and $r_s$.
%
The dark matter density profile defined in this way is determined uniquely by the halo mass, the stellar mass and the half-light radius (the more concentrated the stellar distribution, the stronger the halo response, the steeper the dark matter density profile).
%However, we wanted to explore scenarios in which there is scatter in the dark matter inner slope. Therefore, we applied a random scatter on $\gammadm$ of amplitude $\sigma_{\gamma}$. We explored different values for $\sigma_{\gamma}$, as explained in the next section.

@@

Early-type galaxies lie on the stellar mass fundamental plane, a scaling relation between stellar mass, half-light radius and central velocity dispersion \citep{H+B09}:
\begin{equation}\label{eq:fpcompact}
\sigma_{\mathrm{e}} \propto \mstar^{(\mathrm{obs})\beta_{\sigma}}\reff^{\xi_{\sigma}}.
\end{equation}
The existence of a fundamental plane relation is a consequence of the virial theorem: the velocity dispersion of a galaxy is directly related to the 3-dimensional mass distribution in its inner regions, which is typically dominated by the stellar component.
At fixed observed stellar mass distribution, however, the central velocity dispersion can vary depending on the stellar population synthesis mismatch parameter, on the dark matter mass and on the dark matter density profile. This gives rise to a spread in the values of the velocity dispersion given $\mobs$ and $\reff$.
Therefore, we can use the observed scatter in velocity dispersion to put an upper limit on the intrinsic scatter in the stellar population synthesis mismatch parameter, halo mass and dark matter slope.

@@

For the simulation with extended sources, we produced images with a pixel size of $0.1''$.
We applied a Moffat PSF with an FWHM of $0.2''$.
Finally, we assumed a background noise level such that an extended source with half-light radius $0.5''$ and an apparent magnitude in the absence of lensing of $\msource=25$ is detected with $S/N=10$.
These values are close to those expected for the Euclid Wide survey \citep{Sca++22}.
%
We then applied the peak detection-based lens selection criterion introduced in section \ref{ssec:lensdefext} to find the strong lenses.

@@

\subsection{Limitations of our analysis}

@@

>>>>

We fix the S\'{e}rsic index of the lenses to $n=4$, corresponding to a de Vaucouleurs model.
Although early-type galaxies are often described with a free S\'{e}rsic index, a de Vaucouleurs profile is able to reproduce their surface brightness distribution to a few percent in the radial range $1\rm{kpc} < R < 30\rm{kpc}$ \citep[see e.g.][]{Son20}, which is the region that is most relevant for strong lensing.
Finally, we assume that the light distribution of the stellar component follows its mass distribution exactly. That is, we do not allow for the presence of gradients in the stellar mass-to-light ratio. This assumption is probably broken in real galaxies but, as we discuss in \Sref{ssec:limitations}, does not impact our investigation.

@@

This result suggests that, to first approximation, lensing does not introduce a strong selection in source size.
This follows from the fact that gravitational lensing preserves surface brightness: a source that can be detected in the absence of lensing will produce images with the same surface brightness when lensed, which can be detected as well.
%When the source is extended, however, 
In order to classify a source as strongly lensed, however, we require that multiple images are observed. 
Only sources that lie within a well-defined region give rise to a strong lensing configuration.
If part of the source extends outside of this region, then only a fraction of its flux contributes to creating a set of strongly lensed images.
This lowers the signal-to-noise ratio of the multiple images compared to the point source case.
%The result is the decrease in $\crosssect$ at values of the source size smaller than the 
The result is that $\crosssect$ starts to decrease with increasing source size at values of $\theta_{\mathrm{e,s}}$ that are smaller than the no-lensing detection limit, as observed in \Fref{fig:extcs}.

@@

To determine the density profile of the halo we relied on a theoretically motivated model that takes into account the effect of baryons on the dark matter.
We assumed the dark matter distribution to be initially described by an NFW profile with a concentration\footnote{The concentration is the ratio between the halo virial radius and the scale radius $r_s$.} of five, then used the prescription of \citet{Cau++20} to model the response of the halo to the infall of baryons. % on the basis of the present stellar mass distribution. 
In the \citet{Cau++20} model the halo response is approximated with an analytical function that depends on the present stellar mass distribution, and typically results in a more concentrated and steeper density profile compared to the original NFW model. 
Finally, we fitted a gNFW profile to the surface mass density of the contracted halo. By doing this, we were able to fully describe the dark matter density profile with three parameters: $\mhalo$, $\gammadm$ and $r_s$.
%
The dark matter density profile defined in this way is determined uniquely by the halo mass, the stellar mass and the half-light radius (the more concentrated the stellar distribution, the stronger the halo response and the steeper the dark matter density profile).
In principle we could have allowed for additional degrees of freedom, for instance by relaxing the assumption of a fixed initial halo concentration.
In practice, as we explain in section \ref{ssec:limitations}, our main resultsare not affected by this choice.
%However, we wanted to explore scenarios in which there is scatter in the dark matter inner slope. Therefore, we applied a random scatter on $\gammadm$ of amplitude $\sigma_{\gamma}$. We explored different values for $\sigma_{\gamma}$, as explained in the next section.

@@

Early-type galaxies lie on the stellar mass fundamental plane, a scaling relation between stellar mass, half-light radius and central velocity dispersion \citep{H+B09,deG++21}:
\begin{equation}\label{eq:fpcompact}
\sigma_{\mathrm{e}} \propto \mstar^{(\mathrm{obs})\beta_{\sigma}}\reff^{\xi_{\sigma}}.
\end{equation}
The existence of a fundamental plane relation is a consequence of the virial theorem: the velocity dispersion of a galaxy in dynamical equilibrium is directly related to the 3-dimensional mass distribution in its inner regions, which is typically dominated by the stellar component.
At fixed observed stellar mass distribution, however, the central velocity dispersion can vary depending on the stellar population synthesis mismatch parameter, on the dark matter mass and on the dark matter density profile. This gives rise to a spread in the values of the velocity dispersion given $\mobs$ and $\reff$.
Therefore, we can use the observed scatter in velocity dispersion to put an upper limit on the intrinsic scatter in the stellar population synthesis mismatch parameter, halo mass and dark matter slope.

@@

For the simulation with extended sources, we produced images with properties similar to those expected for the Euclid Wide survey 
citep{Sca++22}.
We used a pixel size of $0.1''$ and we applied a Moffat PSF with an FWHM of $0.2''$ and a $\beta$ parameter of $5.0$.
Finally, we assumed a background noise level such that an extended source with half-light radius $0.5''$ and an apparent magnitude in the absence of lensing of $\msource=25$ is detected with $S/N=10$.
%These values are close to those expected for the Euclid Wide survey \citep{Sca++22}.
%
We then applied the peak detection-based lens selection criterion introduced in section \ref{ssec:lensdefext} to find the strong lenses.

@@

\subsection{Limitations of our analysis}\label{ssec:limitations}

@@

We also assumed that the dark matter density profile of the galaxies is determined uniquely by the halo mass and by the stellar mass distribution (see section \ref{ssub:dmprofile}). In reality we expect there to be a range of density profiles, for example as a result of halos having a nonzero scatter in their initial concentration parameter.
We could in principle carry out an experiment with such an additional source of variation in the dark matter profile.
Allowing for additional flexibility at fixed halo mass scatter parameter $\sigma_h$ results in a larger scatter in the inner dark matter distribution, which would increase the amplitude of the lensing bias. However, the value of $\sigma_h$ that we chose for our fiducial experiment was tuned on the basis of the predicted scatter in velocity dispersion around the fundamental plane. That scatter would also be increased by adding flexibility to the dark matter density profile. Then, in order to be consistent with the analysis carried out in our paper, the value of $\sigma_h$ would need to be lowered accordingly. This, in turn, would reduce the amplitude of the strong lensing bias.
Since lensing and dynamics are sensitive to the mass distribution at comparable scales (the half-light radius and Einstein radius are similar for most of the lenses), we expect these two effects to cancel out to first approximation, leaving the strong lensing bias roughly unchanged.
